IS 607 Assignment 4
========================================================

In order to calculate the 'best popular' year for movies I focused on the 
rating and votes fields. I took rating as a proxy for best and votes as a
proxy for popularity.

I used quartiles to break the votes into 4 groups and created a new variable 
called 'qvotes' - with 1 for the number of votes below the lower quartile up to 4 for the number of votes above the upper quartile.

Then I created a function to create the 'best popular' combination. For this
I simply summed the rating to the 'qvote', to get a 'best popular' measure.

Using ddply I built a dataframe containing the 'best popular' measure for each
year. Then, using this new dataframe, I used a scatter plot to show the 'best 
popular' measure for each year.

```{r}
require(ggplot2)
require(plyr)

#Download the data
#getwd()
df1 <- read.table("movies.tab", sep="\t", header=TRUE, quote="", comment="")
#df1 <- read.table("movies.small.tab", sep="\t", header=TRUE, quote="", comment="")
head(df1)
tail(df1)
#length(df1$year)
#str(df1)
```

Add extra column (qvotes) to df1 based on the quartiles for the number of votes. 

This extra field contain a number from 1-4:

1 - if the number of votes less than or equal to 25th quartile

2 - if the number of votes between the 25th quartile & 50th quartile

3 - if the number of votes between the 50th quartile & 75th quartile

4 - if the number of votes between is greater than the 75th quartile

Create a function (quant.fun) to calculate the appropriate 1-4 value.
Use the quantile() function and some loops to populate the quant.fun output.

```{r}
quant.fun <- function(input)
{
  quant.vec <- quantile(input$votes)
  input.vec <- input$votes
  output.vec <- rep(NA,length(input.vec))
  for(j in 1:length(input.vec)){
    if(input.vec[j]<=quant.vec[2]){
      output.vec[j] <- 1
    } else if (input.vec[j]<=quant.vec[3]){
      output.vec[j] <- 2
    } else if (input.vec[j]<=quant.vec[4]){
      output.vec[j] <- 3
    } else if (input.vec[j]<=quant.vec[5]){
      output.vec[j] <- 4
    }
  }  
  output.vec
}
```

```{r}
#Run quant.fun
df1$qvotes <- quant.fun(df1)
#check to see that new column (qvotes) was added as expected
head(df1)
```

```{r}
#Write a function (bestpop) to calculate the 'best popular' year based on rating and qvotes; In order to capture the quantity of votes (popularity) along with the rating (best), add the qvote to the rating and get the average of this value per year

bestpop <- function(input)
{
  c(BP = with(input, mean(rating+qvotes)))
}
```

```{r}
bp.year <- ddply(df1, .variables="year", .fun="bestpop")
#head(bp.year)
#rank bp.year to see the 'best popular' years
head(bp.year[with(bp.year, order(-bestpop)), ])
```

```{r fig.width=7, fig.height=6}
#plot the 'best popular' number per year
ggplot(bp.year, aes(x=year, y=bestpop)) +geom_point(shape=1) + geom_smooth(method=lm)
```

```{r}
#print out the 'best popular' film year
print(paste("The 'best popular' film year is: ", bp.year[with(bp.year, order(-bestpop)), ][1,1]))
```

Conclusions

It is quite suspicious that the 'best popular' film in from 1893, where only one film was rated for the whole year.

In fact, the 'best popular' values before ~1930 are much further away from the
trend line. 

To investigate this I looked into the frequency of observations per year.

```{r}
#check to see the frequency of observations per year
df.films <- aggregate(title ~ year, df1, length)
head(df.films)
```

```{r fig.width=7, fig.height=6}
ggplot(df.films, aes(x=year, y=title)) +geom_point(shape=1) + geom_smooth(method=lm)
```

As clearly seen in the plot above, there are not many observations pre 1930. Therefore, here is a revised 'best popular' film year (post 1930)!

```{r}
print(paste("The 'best popular' film year (post 1930) is: ", bp.year[with(bp.year, order(-subset(bestpop,year>1930))), ][1,1]))
```

Appendix

Some extra code I wrote to do more simplistic ranking

```{r}
#average rating per year
df.rating <- aggregate(rating ~ year, df1, mean)
#head(df.rating)
```

```{r fig.width=7, fig.height=6}
#scatter plot with regression line of avg rating per year
ggplot(df.rating, aes(x=year, y=rating)) +geom_point(shape=1) + geom_smooth(method=lm)
```

```{r}
#sort df.rating by avg rating in descending order
head(df.rating[with(df.rating, order(-rating)), ])
```

```{r}
#average votes per year
df.votes <- aggregate(votes ~ year, df1, mean)
#head(df.votes)
```

```{r fig.width=7, fig.height=6}
#scatter plot with regression line of avg votes per year
ggplot(df.votes, aes(x=year, y=votes)) +geom_point(shape=1) + geom_smooth(method=lm)
```

```{r}
#sort df.votes by avg votes in descending order
head(df.votes[with(df.votes, order(-votes)), ])
```

```{r}
#examine the relationship between votes and rating
#join df.rating with df.votes on year using plyr join function
df.ratingvotes <- join(x=df.rating, y=df.votes, by=c("year"))
```

```{r fig.width=7, fig.height=6}
#plot the relationship between avg rating & avg votes per year
ggplot(df.ratingvotes, aes(x=rating, y=votes)) +geom_point(shape=1) + geom_smooth(method=lm)
```

```{r fig.width=7, fig.height=6}
#plot the relationship between film rating & film votes
ggplot(df1, aes(x=rating, y=votes)) +geom_point(shape=1) + geom_smooth(method=lm)
```